Naturalness and Long-lived Beyond the Standard Model Particles

### 2017-07-27 tex/draft.tex
Update SM to reflect Andrew's comments

\section{The Standard Model}
    The standard model of particle physics (SM) describes all known particles and their non-gravitational interactions and is experimentally verified to astonishing \fxnote{weak word} accuracy. The SM consists of 12 spin-1/2 particles, the fermions, that make up all observed matter; 12 spin-1 particles, the gauge bosons, that communicate the electromagnetic, weak, and strong forces; and 1 fundamental scalar, the Higgs boson, which will be described in detail below.

    All SM particles transform according to the gauged symmetry group SU(3)cxSU(2)LxU(1)Y, where SU(3)c is associated with color charge, SU(2)L is associated with weak isospin, and U(1)Y is associated with weak hypercharge. The gauge bosons are associated with the generators of these groups, thus there are 8 gluons associated with SU(3)c, three W bosons associated with SU(2)L, and 1 B boson associated with U(1)Y. The fermions are catagorized by which representation of each group they fit into. For example, SU(3)c distinguishes the quarks (which live in the fundamental representation) from the leptons (which are singlets), and SU(2)L distinguishes the left-handed fermions (doublets) from the right-handed fermions (singlets). Finally, different values of weak hypercharge further distignuish SM particles according to the Gell-Mann\fxnote{em-dash}Nishijima formula, $iY=2(Q-I3$.

    The Higgs potential has a nonzero vacuum expectation value (VEV), which spontaneously breaks SU(2)LxU(1)Y to U(1)EM while leaving SU(3)c unscathed. Once the dust settles, the B and 3 W bosons have been mixed to produce the W+, W-, Z bosons that mediate the weak force and the photon that mediates electromagnetism. \fxnote{Andrew says the next sentence isn't quite right. 3 of 4 DOF from complex Higgs field are eaten by W+,W-,Z. Remaining DOF becomes physical Higgs boson. Masses don't come from broken generators. Read about it then fix} 3 of the 4 generators are broken in the process, leaving 3 extra degrees of freedom that ultimately become the masses of the W+, W-, and Z bosons while the remaining symmetry is associated with electric charge. Furthermore, expanding the Higgs field about its VEV produces mass terms for the quarks and charged leptons, thus giving mass to all massive SM particles.\fxnote{Fermion masses ad-hoc, not fundamental, says Andrew}  \fxnote{probably show a little math}\fxnote{Why can't we add explicit mass terms to SM lagrangian w/o Higgs? Summarize section 1 of Andrew's paper}

    \fxnote{kinda funny wording throughout. Maybe it would be better to trade some words for math?}
    All SM particles interact with virtual particles through loop interactions. These interactions lead to corrections to SM parameters such as particle masses. It is common to write a given parameter as a linear combination of its 'bare' (uncorrected) value and its corrections.\fxnote{Not wrong, but sounds funny. Of course we write it that way when talking about a sum of diagrams} As the only fundamental scalar in the SM, the Higgs is uniquely affected by these corrections.\fxnote{Explain why scalars are more sensitive. Show math probably} The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the Higgs bare mass and the terms that correct it.

\section{Naturalness and the Higgs mass}
    \fxnote{cite guidice naturally and t'hooft naturalness}
    \fxnote{Haven't specifically said 'hierarchy problem. Is that bad?}
    \fxnote{Never really talked about the effective theory concept or dimensional analysis}
    The naturalness criterion can be stated in many ways, a few of which are reproduced here: (1) All dimensionless parameters should be of order 1, (2) An effective theory should not be sensitive to small variations in the parameters of higher-energy theory, and (3) any dimensionless parameter much smaller than 1 must be protected by a custodial symmetry. When applied to the Higgs mass, naturalness leads to what's known as the 'hierarchy problem' or unacceptable levels of 'fine tuning'. These issues are unpacked below.

@@

    \fxnote{Point is that corrections come in with different signs and individual terms are huge. Really weird that they would cancel out so perfectly. Maybe add feynman diagrams to illustrate. Use Jaxodraw probably} Order($10^{32}$) corrections imply that the bare Higgs mass must be negative and also order($10^{32}$) to reproduce the experimetnally known value. In fact, the bare mass and corrections must match to 1 part in $\approx 10^{30}$. While it's entirely possible that these two numbers just so happen to match so perfectly, many \fxnote{weak} feel that such miraculous fine tuning must be a hint that some deeper physical mechanism is at work. That is, either some previously unknown symmetry is protecting the Higgs mass or the cutoff scale is order(TeV).

@@

\section{Supersymmetry}
    Supersymmetry (SUSY) introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between bosons and fermions. In its simplest form, SUSY results in one new boson for every SM fermion and one new fermion for every SM boson \fxnote{and an extra Higgs field - probably should mention it. Make sure I understand it either way}. The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-1/2 superpartners to the SM bosons are given the name of their SM counterpart with "ino" tacked on to the end (e.g. Higgsino or wino). Together the superpartners to the SM particles are known as sparticles.

@@

    After estimating the background by extrapolating data from the cosmic-enriched region to the signal region and determining systematic uncertainties related to the MC simulation, trigger acceptance, and luminosity, the authors calculated the expected background for LLP lifetimes ranging from 100 ns to 1 us. In all cases, the expected background was less then one event. Exactly zero events are observed in every lifetime scenario. These results are then interpreted as upper limits on the signal production cross section \fxnote{include limit plots or table}, which are the first limits for stopped particles that decay to muons at the LHC.

>>>>

\section{The Standard Model}
    The standard model of particle physics (SM) describes all known particles and their non-gravitational interactions and is experimentally verified to astonishing \fxnote{weak word} accuracy. The SM consists of 12 spin-$\frac{1}{2}$ particles, the fermions, that make up all observed matter; 12 spin-1 particles, the gauge bosons, that communicate the electromagnetic, weak, and strong forces; and 1 fundamental scalar, the Higgs boson, which will be described in detail below.

    All SM particles transform according to the gauged symmetry group $\SUthreecSUtwoLUoneY$, where $\SUthreec$ is associated with color charge, $\SUtwoL$ is associated with weak isospin, and $\UoneY$ is associated with weak hypercharge. The gauge bosons are associated with the generators of these groups, thus there are 8 gluons associated with $\SUthreec$, three W bosons associated with $\SUtwoL$, and one B boson associated with $\UoneY$. The physical gauge bosons, which are linear combinations of the Ws and B, are the $\Wp$, $\Wm$, $\Z$, and photon ($\gamma$). The fermions are catagorized by which representation of each group they fit into. For example, $\SUthreec$ distinguishes the quarks (which live in the fundamental representation) from the leptons (which are singlets), and $\SUtwoL$ distinguishes the left-handed fermions (doublets) from the right-handed fermions (singlets). Finally, weak hypercharge further distignuishes SM particles according to the Gell-Mann--Nishijima formula, $Y=Q-I_3$, where Q is electric charge and $I_3$ is the third component of weak isospin.

    Experimentally, the $\Wpm$, $\Z$, quarks, and charged leptons are massive\fxnote{Gloss over neutrino mass?}. Explicit mass terms for the fermions or gauge bosons would violate the chiral and gauge symmetries of the SM Lagrangian, however, so $\SUtwoLUoneY$ must be broken. Specifically, the nonzero vacuum expectation value (VEV) of the Higgs field spontaneously breaks $\SUtwoLUoneY$ to $\UoneQ$. In the process, three of the four degrees of freedom associated with the complex Higgs doublet become the longitudinal polarizations of $\Wpm$ and $\Z$, giving them mass. The remaining degree of freedom is the physical Higgs boson. Expanding the Higgs field about its VEV produces Yukawa interactions such as those shown below. These terms are interpreted as giving the quarks and charged leptons mass, where the mass of each particle is given by the product of the Higgs VEV and the coupling constant.

    \fxnote{Yukawa terms go here}

    SM parameters, such as particle masses\fxnote{what else is corrected. How?}, are corrected by interactions with virtual particles through loop diagrams. Figure\fxnote{make and reference diagrams}, for example, shows the largest SM corrections to the Higgs and electron masses. Because the Higgs is a fundamental scalar, its corrections depend quadratically on the cutoff scale, $\Lambda$, which is the largest energy scale at which the SM is valid. The fermions and gauge bosons, on the other hand, only depend logarithmically\fxnote{check. be able to work out example} on $\Lambda$. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the bare Higgs mass and the terms that correct it.

    \fxnote{Feynman diagrams go here}

\section{Naturalness and the Higgs mass}
    \fxnote{cite guidice naturally and t'hooft naturalness}
    \fxnote{Haven't specifically said 'hierarchy problem. Is that bad?}
    \fxnote{Never really talked about the effective theory concept or dimensional analysis}
    The naturalness criterion can be stated in many ways, a few of which are reproduced here: (1) All dimensionless parameters should be of order one, (2) An effective theory should not be sensitive to small variations in the parameters of higher-energy theory, and (3) any dimensionless parameter much smaller than one must be protected by a custodial symmetry. When applied to the Higgs mass, naturalness leads to what's known as the `hierarchy problem' or unacceptable levels of `fine tuning'. These issues are unpacked below.

@@

    \fxnote{Point is that corrections come in with different signs and individual terms are huge. Really weird that they would cancel out so perfectly. Maybe add feynman diagrams to illustrate. Use Jaxodraw probably} Order($10^{32}$) corrections imply that the bare Higgs mass must be negative and also order($10^{32}$) to reproduce the experimetnally known value. In fact, the bare mass and corrections must match to one part in $\approx 10^{30}$. While it's entirely possible that these two numbers just so happen to match so perfectly, many \fxnote{weak} feel that such miraculous fine tuning must be a hint that some deeper physical mechanism is at work. That is, either some previously unknown symmetry is protecting the Higgs mass or the cutoff scale is order(TeV).

@@

\section{Supersymmetry}
    Supersymmetry (SUSY) introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between bosons and fermions. In its simplest form, SUSY results in one new boson for every SM fermion and one new fermion for every SM boson \fxnote{and an extra Higgs field - probably should mention it. Make sure I understand it either way}. The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-$\frac{1}{2}$ superpartners to the SM bosons are given the name of their SM counterpart with "ino" tacked on to the end (e.g. Higgsino or wino). Together the superpartners to the SM particles are known as sparticles.

@@

    After estimating the background by extrapolating data from the cosmic-enriched region to the signal region and determining systematic uncertainties related to the MC simulation, trigger acceptance, and luminosity, the authors calculated the expected background for LLP lifetimes ranging from 100 ns to 1 us. In all cases, the expected background was less then one event. Exactly zero events are observed in every lifetime scenario. These results are then interpreted as upper limits on the signal production cross section\fxnote{and mass limits}\fxnote{include limit plots or table}, which are the first limits for stopped particles that decay to muons at the LHC.
